Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors

### 2021-10-08 DOI_Review/Manuscript.rmd
Text edits mostly

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: Conservation researchers are increasingly identifying diffusion theory as useful for understanding the spread of community-based conservation initiatives. This work has nevertheless been largely ambiguous as to the specific opportunities and shortcomings for diffusion theory in conservation research, and how it might be practically applied. Here, we employ agent-based simulations to confirm that process-based diffusion models reasonably predict the long-term patterns in adoption of conservation behaviors even under tightly coupled behavior-environment feedbacks. We caution however, that while diffusion theory may predict the pattern with which a conservation intervention will spread through a population, it gives little insight as to the underlying transmission mechanism. We show that the transmission mechanism employed by a population can alter the final proportion of individuals that adopt an intervention, and argue that conservation researchers and practitioners can leverage theory from social learning and cultural evolution to better design and diagnose projects. Lastly, we emphasize that developing and fitting process models in statistical frameworks traditionally reserved for regression alone, may pose a productive new avenue for conservation science, particularly in our understanding of conservation as a complex system.

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral uptake [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. In this instantiation, $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} \\
&\frac{dR}{dt} = \gamma* \text{I}_{t} - \chi * \text{R}_{t}
\end{align*}

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, &  SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017;  @carpenter_predator-prey_2018]. Below, we use a minimal agent-based model of conservation adoption contingent on behavior-environment feedbacks to simulate data on the long-term adoption dynamics of a community-based conservation intervention. Then, using just the first 30% of the data, we estimate the rate of intervention spread, rate of dropout, and time after dropout before agents are available to adopt again. We use these estimated parameters to predict the overall patterns of adoption over all time steps and compare our observed to predicted values.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under both unbiased and success-biased social learning. Success-biased social learning is contingent on behavior-environment feedbacks. Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (from herein referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first timestep, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and individuals are rewarded for abstaining from harvesting there. In each subsequent timestep, agents either choose to enroll in the conservation program, receive their reward, and continue to harvest just from the remaining 80% of the unprotected landscape, or they choose to harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior [@baldini_success-biased_2012; @baldini_two_2013, @barrett_equifinality_2019]. The protected and unprotected resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area.

@@

$$ \text{Y}_{t} \sim Binomial(\text{N}, \text{I}_{t}) $$
\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} \\
&\frac{dR}{dt} = \gamma* \text{I}_{t} - \chi * \text{R}_{t}
\end{align*}

@@

\begin{align*}
&\frac{dS}{dt} = - \beta * \text{S}_{t} * \text{I}_{t} + \chi * \text{R}_{t} \\
&\frac{dI}{dt} = \beta * \text{S}_{t} * \text{I}_{t} - \gamma * \text{I}_{t} - \phi * \text{E}_{t} \\
&\frac{dR}{dt} = \gamma * \text{I}_{t} - \chi * \text{R}_{t} + \phi * \text{E}_{t} \\
&\frac{dE}{dt} =  - \rho * (\text{S}_{t} + \text{R}_{t}) + \theta*\text{E}_{t}
\end{align*}

>>>>

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation researchers and practitioners is to design interventions that benefit both human and ecological well-being and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding the spread of these community-based conservation initiatives. This work has nevertheless been largely ambiguous as to the practical applications of diffusion theory to community-based conservation and has not addressed the potential shortcomings for its application to complex, coupled human-natural systems. Here, we employ agent-based simulations to confirm that process-based diffusion models reasonably predict the long-term patterns in adoption of conservation behaviors even under tightly coupled behavior-environment feedbacks. We caution however, that while diffusion theory may predict the pattern with which a conservation intervention will spread through a population, it gives little insight as to the underlying transmission mechanism. We show that the transmission mechanism employed by a population can alter the final proportion of individuals that adopt an intervention, and argue that conservation researchers and practitioners can leverage theory from social learning and cultural evolution to better design and diagnose projects. Lastly, we emphasize that developing and fitting process models in statistical frameworks traditionally reserved for regression alone, may pose a productive new avenue for conservation science, particularly in concert with simulation techniques and insights from complex systems.

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral uptake [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. These compartmental differential equations assign every member of a population to a category (S, I, or R), depending on the proportion of individuals in the other categories and parameter values for the natural history characteristics of a given disease [@tolles_modeling_2020]. In this instantiation, $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align*}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t}
\end{align*}

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, &  SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018].

Below, we develop a minimal agent-based model to highlight our core dynamics of interest. In this abstraction, we simulate conservation adoption contingent on behavior-environment feedbacks to produce data on the long-term adoption dynamics of a community-based conservation intervention. Then, using just the first 30% of the data, we estimate the rate of intervention spread, rate of dropout, and readoption. We use these estimated parameters to predict the overall patterns of adoption over all time steps and compare our observed to predicted values.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under two social learning mechanisms. The first is unbiased copying, where agents have the chance to copy an alternative behavior with some set probability independent of the benefit provided by adopting the behavior or not [@barrett_equifinality_2019]. The second mechanism is success-biased learning, under which, agents disproportionately copy the behaviors of more successful agents [@baldini_success-biased_2012]. Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (from herein referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first timestep, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and individuals are rewarded for abstaining from harvesting there. In each subsequent timestep, agents either choose to enroll in the conservation program, receive their reward, and continue to harvest just from the remaining 80% of the unprotected landscape, or they choose to harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior [@baldini_two_2013]. The protected and unprotected resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area. As behavioral adoption under success-biased learning is dependent on harvest success, the successfulness of each behavior is determined largely by the number of available resources in each area, and the number of resources in each area is a result of behavioral adoption, this learning strategy is contingent on behavior-environment feedbacks, unlike unbiased adoption which occurs independent of harvest success.

@@

$$ \text{Y}_{t} \sim Binomial(\text{N}, \text{I}_{t}) $$
\begin{align*}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t}
\end{align*}

@@

\begin{align*}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} - \phi \cdot \text{E}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t} + \phi \cdot \text{E}_{t} \\
&\frac{dE}{dt} =  - \rho \cdot (\text{S}_{t} + \text{R}_{t}) + \theta \cdot \text{E}_{t}
\end{align*}
